---
title: "Geog4/6300: Lab 1"
format: gfm
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

## Loading data into R, data transformation, and summary statistics

**Your name: Grayson Smith**

**Overview and lab criteria:**

This lab is intended to assess your ability to use R to load data and to generate basic descriptive statistics. For this lab to be marked complete, the following criteria must be met:

4. Identify and apply appropriate data filtering and cleaning strategies to prepare datasets for analysis. (Task 2)
5. Effectively interpret the code you create, explaining in plain language what each step does to the data. (Task 6)
6. Identify and use appropriate external documentation — including package references, help files, and peer resources — to learn and apply unfamiliar functions or methods. (Task 7)
7. Filter, aggregate, and transform datasets using grouping and summary operations to answer specific analytical questions. (Tasks 1 & 5)
9. Reshape data between wide and long formats to meet the requirements of different analytical or visualization tasks. (Task 4)
11. Create effective data visualizations across multiple chart types (line, scatter, histogram, Q-Q plot), applying appropriate aesthetic choices such as color, grouping, and labeling. (Task 3 & 5)

**Data:**

You'll be using monthly weather data from the Daymet climate database (http://daymet.ornl.gov) for all counties in the United States over an 12-year period (2010-2021). These data are available on the GitHub repo for our course. The following variables are provided:

* `cty_txt`: Code for joining to census data
* `year`: Year of observation (with an initial "Y" to make it a character)
* `month`: Month of observation (1 = Jan, 2 = Feb, etc.)
* `median_tmax`: Median maximum recorded temperature (Celsius)
* `median_tmin`: Median minimum recorded temperature (Celsius)
* `sum_prcp`: Total recorded precipitation for the month (mm)
* `cty_name`: Name of the county
* `state`: State of the county
* `region`: Census region (map: https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf)
* `division`: Census division
* `X`: Longitude of the county centroid
* `Y`: Latitude of the county centroid

These labs are meant to be done collaboratively, but your final submission should demonstrate your own original thought (don't just copy your classmate's work or turn in identical assignments). Your answers to the lab questions should be typed in this Quarto template. You'll then render the document to a GitHub markdown document and upload it to your class GitHub repo.

**Procedure:**

Load the tidyverse package and import the data:

```{r}
#| message: false

library(tidyverse)

daymet_data <- read_csv("data/daymet_monthly_median_2010-2021.csv")
```

We can look at the first few rows of the dataset using the *head()* function. We also use *kable* to format the output as a readable table..

```{r}
kable(head(daymet_data))
```

There are a lot of observations here, 452,448 to be exact. To get a better grasp on the data, we can use `group_by()` and `summarise()` from the tidyverse package. This will allow us to identify the mean value for each year by county across the study period.

## Task 1

*Use `group_by()` and `summarise()` to calculate the mean minimum temperature for each year by county across all months, also including State and Region as grouping variables. Your resulting dataset should show the value of tmin for each county in each year. Use the `kable()` and `head()` functions as shown above to call the resulting table.*

```{r}
#| label: task1
# Your code goes here

grouped_daymet_data <- daymet_data %>%
  group_by(cty_name, state, region, year) %>% 
  summarise(mean_tmin_bycty = mean(median_tmin)) %>%
  mutate(year = str_remove(year,"Y"))
kable(grouped_daymet_data)
head(grouped_daymet_data)
```

## Task 2

*Let's shift to the state level, focusing on those in the South Region. Filter the original data frame (`daymet_data`) to just include counties in this region. Then calculate the mean minimum temperature by year for each state. For an optional extra challenge, use the `round()` function to include only 1 decimal point. Use `kable()` and `head()` to call the first few lines of the resulting table.*

```{r}
#| label: task2

# Your code goes here

grouped_daymet_data_south <- daymet_data %>%
  filter(region=="South Region") %>%
  group_by(state, year) %>% 
  summarise(mean_tmin_bystate = mean(median_tmin)) %>%
  mutate(rounded_mean_tmin_bystate = round(mean_tmin_bystate, digits = 1)) %>%
  mutate(year = str_remove(year,"Y"))
kable(grouped_daymet_data_south)
head(grouped_daymet_data_south)
```

## Task 3

*To visualize the trends, we could use ggplot to visualize change in mean temperature over time. Create a line plot (`geom_line`) showing the state means you calculated in task 2. Use the `color` parameter to show separate colors for each state. You may also need to define the state as a group in the aesthetic parameter.*

```{r}
#| label: task3

# Your code goes here

ggplot(grouped_daymet_data_south,
       aes(x=year, 
           y=mean_tmin_bystate, group=state,
           color=state)) +
  geom_line() +
  ggtitle("Annual Mean Min Temp for Southern States (2010-2021)") +
  theme(
    plot.title = element_text(size = 12)) + #Had to do to shrink title to fit
  xlab("Year") +
  ylab("Mean Minimum Temp (\u00B0C)")
```

## Task 4

*If you wanted to look at these data as a table, you'd need to have it in wide format. Use the `pivot_wider()` function to create a wide-format version of the data frame you created in task 2. In this case, the rows should be states, the columns should be the years, and the data in those columns should be mean minimum temperatures. Then call the whole table using `kable()`.*

```{r}
#| label: task4

# Your code goes here
wider_grouped_daymet_data_south <- grouped_daymet_data_south %>%
  group_by(state) %>%
  pivot_wider(id_cols = state, names_from= year,
              values_from= rounded_mean_tmin_bystate) #The id_cols groups everything so we have one state per row; without it, you get one row for each State-Year combo. 


kable(wider_grouped_daymet_data_south)
```

## Task 5

*Let's assess the relationship of heat and precipitation by region. Returning to the original dataset, create a data frame that shows the mean maximum temperature and mean precipitation for all states in 2015, also including region as a subgroup in your `group_by`. Then use ggplot to create a scatterplot (`geom_point`) for these two variables, coloring the points using the region variable.*

```{r}
#| label: task5

# Your code goes here
grouped_daymet_data_region_2015 <- daymet_data %>%
  filter(year == "Y2015") %>%
  group_by(state, region, year) %>% 
  summarise(mean_tmax_bystate = mean(median_tmax), mean_sumprcp_bystate = mean(sum_prcp)) %>%
  mutate(year = str_remove(year,"Y"))
#kable(grouped_daymet_data_region)

#Making the plot itself
ggplot(grouped_daymet_data_region_2015,
       aes(x=mean_sumprcp_bystate, 
           y=mean_tmax_bystate,
           color=region)) +
  geom_point() +
  ggtitle("Annual Mean Max Temp and Precip by State and Region (2015)") +
  theme(
    plot.title = element_text(size = 11)) +
  xlab("Mean Precip (inches)") +
  ylab("Mean Maximum Temp (\u00B0C)") +
  labs(color = "Region") #Had to do this to capitilize "region" in the legend. 

```

## Task 6

*In the space below, explain what each function in your code for task 5 does to the dataset in plain English.*

Task 6 Response: The first part of Task 5 -- the data manipulation portion -- begins by creating a new variable (grouped_daymet_data_region_2015). From there, the original dataset (daymet_data) is then passed through a series of functions, starting with a filter that only takes out rows for the year 2015. From there, the data is then grouped according to its state, region, and year (in that order). Using those groupings, the function then takes the mean of the median_tmax and total precipitation for each state in 2015 (the "per state" nature of the data is likely a result of state being the first subgroup by which the data is grouped) and displays those mean values as new columns (with the titles mean_tmax_bystate and mean_sumprcp_bystate). Finally, the data is manipulated to remove the "Y" in front of all the year numbers, just to allow for a more appealing plot (as described below). 

The code then turns to making a scatter plot using these new mean values for median_tmax and sumprcp (i.e., the mean of the median max temperature and summed precip per state, in 2015). The function starts by referencing the grouped_daymet_data_region_2015 dataset (the data we just manipulated, as described above), then assigns the x and y axes to the mean precip and mean median max temps, respectively. It then colors the respective by-state value for precip and mean max temp (i.e., the point at the "coordinate" pair formed between the precip and temp data for each state) by region (allowing us to see many clear regional trends in temp and precip distribution). From there, the remainder of the code aims to add various titles to the plot itself, its axes, and the legend, all to make the figure more usable and visually appealing. 

## Task 7

*The `dplyr` package also includes `across` function. Use `?across` on the R command line to open the documentation for this function. In the space below, explain what it does in your own words. Then interpret the way the across function is used below, going line by line within the function.*

```{r}
state_2015 <- daymet_data %>%
  filter(year == "Y2015") %>%
  group_by(region, state) %>%
  summarise(
    across(
      c(median_tmax, sum_prcp),
      mean,
      na.rm = TRUE,
      .names = "mean_{.col}"
    )
  )
```

Task 7 Response: It appears that the `across` function can be used to apply another function -- like `summarise` function -- across multiple columns. Using this `across` function should preclude having to do multiple `summarise` functions for different columns -- or at the very least, make the syntax within one`summarise` function more streamlined (as it appears it also possible to apply `summarise` to multiple columns without `across,` it is just far longer syntax). In the code given above, the `across` is used to first take the mean of the values in the median_tmax and sum_prcp, then to remove any N/A values from both of those columns, and finally to name the resultant columns (with the mean values) according to the convention of mean_"previous column name" (i.e., mean_median_tmax). The use of across here just make ts the synatx more straightforward to handle each of those tasks across these two columns without needing two `summarise` functions (or a more complex single `summarise`). 


## Challenge Question

In class, we covered ways of working with the Daymet API. Create a script below that uses the **daymetr** package to download data from Daymet for a place (or places) of your choosing. Then visualize the temporal pattern for a variable of your choosing in this place, similar to what you did in question 4. Use a dplyr function (`mutate()`, `summarise()`, `filter()`, etc.) to do any needed data wrangling and create a visual using ggplot.

In addition to this code, write a short summary of a pattern that's evident in the data you visualized.

```{r}
#| label: challenge

# Your code goes here
library(daymetr)

griffin_daymet <- download_daymet(site="griffin",lat=33.2476,lon=-84.2709,start=2020,end=2025,internal=TRUE) #This is my hometown!
athens_daymet <- download_daymet(site="athens",lat=33.9498,lon=-83.3734,start=2020,end=2025,internal=TRUE)

griffin_df <- as.data.frame(griffin_daymet$data)
athens_df <- as.data.frame(athens_daymet$data)

daymet_data_griffin_yearlytemp_srad <- griffin_df %>%
  filter(tmax..deg.c. > -10, srad..W.m.2. > 0) %>%
  group_by(year) %>%
  summarise(mean_tmax = mean(tmax..deg.c.),mean_srad = mean(srad..W.m.2.)) %>%
  mutate(year = str_remove(year,"Y"),city="Griffin")
#kable(daymet_data_griffin_yearlytemp_srad)

daymet_data_athens_yearlytemp_srad <- athens_df %>%
  filter(tmax..deg.c. > -10, srad..W.m.2. > 0) %>%
  group_by(year) %>%
  summarise(mean_tmax = mean(tmax..deg.c.),mean_srad = mean(srad..W.m.2.)) %>%
  mutate(year = str_remove(year,"Y"),city = "Athens")
#kable(daymet_data_athens_yearlytemp_srad)

combined_griffin_athens_df <- full_join(daymet_data_griffin_yearlytemp_srad,daymet_data_athens_yearlytemp_srad) #Added this because I could not find out a way to plot the data if it was in two separate data frames. That should be investigated for future labs. 

#Making the plot itself
ggplot(combined_griffin_athens_df,
       aes(x=mean_tmax, 
           y=mean_srad, group=city,
           color=city)) +
  geom_point() +
  ggtitle("Annual Mean Temp and Incoming Solar Radiation for Griffin and Athens, GA (2020-2025)") +
  theme(
    plot.title = element_text(size = 9)) + #Had to do to shrink title to fit
  xlab("Mean Max Temp (\u00B0C)") +
  ylab("Mean Solar Radiation (W/m^2)") +
  labs(color = "City")

```

*Explanation goes here.*

## Final Submission Stuff

### Disclosure of Assistance

Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?

The two primary resources I used during the completion of this lab included brief comments from my classmates and some use of Google Gemini. In terms of my interaction with my classmates, I cross-referenced my work (especially for Task 1) with Will to ensure that my answer was reasonable -- allowing me to gain confidence that my strategy for that task was appropriate (and thus allowing me to hopefully remember that strategy for similar tasks in the future [including using Task 1 as a template for much of the group_by and summarise functions done in later lab tasks]). In terms of my use of Google Gemini, I used this tool to get ideas for strategies to troubleshoot my code (without asking for the specific code itself to ensure I wrote the code presented above), especially in areas where the strategies I remembered from class were not producing the reuslt I expected (likely just due to my lacking knowledge). This was especially true in both my inclusion of the "theme" function in many of my plots (a strategy I had to seek out to decrease the size of my overly-long plot titles) and the use of the "id_cols" function in the `pivot_wider`function in Task 4 (which I needed to ensure I got only one result for each state, rather than multiple results [for each year] from the same state). In both cases, Gemini was able to provide advice on what strategies to use to solve my persistent issue, strategies that I then implemented and now intend to use on future tasks. As such, Gemini aided (as with last lab) in deepening my knowledge of different R functionalities -- all of which I hope to retain to solve increasingly complex tasks in future labs. Together, both my collaboration with Will and my use of Gemini aided in my learning here by allowing me to more quickly solve many of my problems while still exposing me to additional R functionalities that I can now use in my future R adventures. 

### Lab Reflection

How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What are the biggest things you learned by completing it?

*Your response here.*